---
title: Flexdashboard in R
author: James Malloy
draft: true
date: '2022-03-18'
slug: []
categories:
  - dashboard
  - data visualization
tags:
  - dashboard
  - data visualization
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)
```

A few months ago, a friend of mine told me about her desire to better track a virtual tutoring program that she manages. I asked her to let me make a dashboard for her, and she agreed. To make the dashboard, I used R and the Flexdashboard package and came up with this as (a replica of) the final product. Below, I'll outline the steps I took to come up with it.

## Step 1:  Access & load data

All the data is input and hosted on Google Sheets, so to access it I used the `googlesheets4` package along with a host of others. This was my first introduction to APIs (application programming interface), which allows two applications to talk to each other (in this case Google and R) and gain access to information that is otherwise private. 

It took me a while, but I was finally able to figure out the API stuff. For this post, I've created "fake" data in Google Sheets and made it publicly viewable. (This simplifies the process). However, I've commented out the code you'd need in real life to access your private Google Sheets data.


```{r load-libraries-data}
library(googlesheets4) # access Google sheets data
library(flexdashboard) # dashboard package, great alternative to Shiny
library(shiny) # I still wanted to use some Shiny functionalities
library(DT) # nice html data tables
library(lubridate) # manipulating date data
library(plotly) # easy-to-code interactive plots
library(janitor) # cleaning names
library(tidyverse) # pipes, data wrangling, etc

# options(
#  gargle_oauth_cache = ".secrets",
#  gargle_oauth_email = TRUE)

# gs4_auth(scope = "https://www.googleapis.com/auth/drive")
# drive_auth(token = gs4_token())

gs4_deauth() # asking Google for permission to access data # gives R permission
             # publicly (not private) viewable Googlesheets only

gs_tutoring <- "1vyVNz2ClxDVcPr7oa1LTc0YyHwZ53DI_lBSvKtl3v8M" # Googlesheet data
```

## Step 2: Tidy the data

Data rarely comes in the form that you need it for proper analysis and visualizations. You often have to manipulate from a "wide" format into a "long" (aka "tidy") format. After your data is in the right format, you transform it to your heart's desire.

{{< panelset class="greetings" >}}
{{< panel name="Not tidy (How it's given) :thumbsdown:" >}}

```{r echo=FALSE}

gs_attendance_col_types <- 
  # define column types c= "character", D = "double"/"dbl" which is just another term for a number
  "ccccccccccccccDDDccccccccccccccccccccccccccccccccccccccccccccnnnn"

df_attendance <-
  read_sheet(gs_tutoring, 
             range = "A9:BM",
             n_max = 100,
             col_types = gs_attendance_col_types)%>% 
  # removed date cols that weren't being used
  select(-c("8/9", "8/16", "8/23", "8/30", "9/6", "9/13", "12/20", "12/27")) %>% 
  mutate(id = row_number(), .before = "Status") %>%
  # changed the day of week values to read out their actual spelling
  mutate(Day = case_when( 
    Day == "MON" ~ "Monday",
    Day == "TUE" ~ "Tuesday",
    Day == "WED" ~ "Wednesday",
    Day == "TH" ~ "Thursday",
    Day == "F" ~ "Friday")) %>% 
  # the dates in the data as it was originally given to me included the "week of"
  # that each tutoring session happened rather than the specific date that the 
  # tutoring happened. Below I create a new variable "day_numeric" that will allow
  # me to calculate the actual date of each tutoring section by doing some simple
  # addition
  mutate(day_numeric = case_when(
    Day == "Monday" ~ 2,
    Day == "Tuesday" ~ 3,
    Day == "Wednesday" ~ 4,
    Day == "Thursday" ~ 5,
    Day == "Friday" ~ 6)) %>%
  mutate(`Attendance Percentage` = round(`Attendance Percentage` , digits = 2) * 100,
         `Attendance Proportion` = `Attendance Percentage` / 100) %>% 
  
  # One day I learn how to do regex :'(
  rename("2021-09-20" = "9/20",
         "2021-09-27" = "9/27",
         "2021-10-04" = "10/4",
         "2021-10-11" = "10/11",
         "2021-10-18" = "10/18",
         "2021-10-25" = "10/25",
         "2021-11-01" = "11/1",
         "2021-11-08" = "11/8",
         "2021-11-15" = "11/15",
         "2021-11-22" = "11/22",
         "2021-11-29" = "11/29",
         "2021-12-06" = "12/6",
         "2021-12-13" = "12/13",
         "2022-01-03" = "1/3",
         "2022-01-10" = "1/10",
         "2022-01-17" = "1/17",
         "2022-01-24" = "1/24",
         "2022-01-31" = "1/31",
         "2022-02-07" = "2/7",
         "2022-02-14" = "2/14",
         "2022-02-21" = "2/21",
         "2022-02-28" = "2/28",
         "2022-03-07" = "3/7",
         "2022-03-14" = "3/14",
         "2022-03-21" = "3/21",
         "2022-03-28" = "3/28",
         "2022-04-04" = "4/4",
         "2022-04-11" = "4/11",
         "2022-04-18" = "4/18",
         "2022-04-25" = "4/25",
         "2022-05-02" = "5/2",
         "2022-05-09" = "5/9",
         "2022-05-16" = "5/16",
         "2022-05-23" = "5/23",
         "2022-05-30" = "5/30",
         "2022-06-06" = "6/6")

# these are the codes used to track attendance at each tutoring session
attendance_codes <- c("X", 1, 2, 3, 4, 5) 

# pivoting the data from wide to low
df_attendance_long <-  df_attendance %>%
  pivot_longer(cols = matches("^[0-9]{4}-[0-9]{2}-[0-9]{2}$"),
               names_to = "week_of",
               values_to = "attendance_code") %>%
  mutate(week_of = ymd(week_of)) %>%
  # sometimes the employees that take attendance mark input "x" or "X", so I 
  # standardize it make all the Xs uppercase
  mutate(attendance_code = case_when(attendance_code == "x" ~ "X", 
                                   TRUE ~ as.character(attendance_code))) %>% 
  # here is where I'm getting a date for each tutoring session rather than
  # than just the week that each session occurred. For example, if a tutoring 
  # session took place during the 2nd day of the week of March 21st, then the 
  # actual date of tutoring will be as follows:
  mutate(tutoring_date = case_when(
    Day == "Monday" ~ as_date(week_of) + 0,    # March 21st + 0 = March 21st
    Day == "Tuesday" ~ as_date(week_of) + 1,   # March 21st + 1 = March 22nd
    Day == "Wednesday" ~ as_date(week_of) + 2, # March 21st + 2 = March 23rd etc
    Day == "Thursday" ~ as_date(week_of) + 3,  # etc, etc. etc.
    Day == "Friday" ~ as_date(week_of) + 4), .before = "Day")%>% 
  relocate(week_of, .after = "Orientation")

df_attendance

```

{{< /panel >}}
{{< panel name="Tidy (How we want it) :thumbsup:" >}}

```{r}
df_attendance_long
```

{{< /panel >}}
{{< panel name="Tidy (How we want it) :thumbsup:" >}}

```{r}
gs_attendance_col_types <-
  "ccccccccccccccDDDccccccccccccccccccccccccccccccccccccccccccccnnnn"

df_attendance <-
  read_sheet(gs_tutoring, 
             range = "A9:BM",
             n_max = 100,
             col_types = gs_attendance_col_types)%>% 
  select(-c("8/9", "8/16", "8/23", "8/30", "9/6", "9/13", "12/20", "12/27")) %>%
  mutate(id = row_number(), .before = "Status") %>%
  mutate(Day = case_when(
    Day == "MON" ~ "Monday",
    Day == "TUE" ~ "Tuesday",
    Day == "WED" ~ "Wednesday",
    Day == "TH" ~ "Thursday",
    Day == "F" ~ "Friday")) %>% 
  mutate(day_numeric = case_when(
    Day == "Monday" ~ 2,
    Day == "Tuesday" ~ 3,
    Day == "Wednesday" ~ 4,
    Day == "Thursday" ~ 5,
    Day == "Friday" ~ 6)) %>%
  mutate(`Attendance Percentage` = round(`Attendance Percentage` , digits = 2) * 100,
         `Attendance Proportion` = `Attendance Percentage` / 100) %>% 
  # What's a better way to do this? Regex?
  rename("2021-09-20" = "9/20",
         "2021-09-27" = "9/27",
         "2021-10-04" = "10/4",
         "2021-10-11" = "10/11",
         "2021-10-18" = "10/18",
         "2021-10-25" = "10/25",
         "2021-11-01" = "11/1",
         "2021-11-08" = "11/8",
         "2021-11-15" = "11/15",
         "2021-11-22" = "11/22",
         "2021-11-29" = "11/29",
         "2021-12-06" = "12/6",
         "2021-12-13" = "12/13",
         "2022-01-03" = "1/3",
         "2022-01-10" = "1/10",
         "2022-01-17" = "1/17",
         "2022-01-24" = "1/24",
         "2022-01-31" = "1/31",
         "2022-02-07" = "2/7",
         "2022-02-14" = "2/14",
         "2022-02-21" = "2/21",
         "2022-02-28" = "2/28",
         "2022-03-07" = "3/7",
         "2022-03-14" = "3/14",
         "2022-03-21" = "3/21",
         "2022-03-28" = "3/28",
         "2022-04-04" = "4/4",
         "2022-04-11" = "4/11",
         "2022-04-18" = "4/18",
         "2022-04-25" = "4/25",
         "2022-05-02" = "5/2",
         "2022-05-09" = "5/9",
         "2022-05-16" = "5/16",
         "2022-05-23" = "5/23",
         "2022-05-30" = "5/30",
         "2022-06-06" = "6/6")

df_attendance_long <-  df_attendance %>%
  pivot_longer(cols = matches("^[0-9]{4}-[0-9]{2}-[0-9]{2}$"),
               names_to = "week_of",
               values_to = "attendance_code") %>%
  mutate(week_of = ymd(week_of)) %>%
  mutate(attendance_code = case_when(attendance_code == "x" ~ "X", 
                                   TRUE ~ as.character(attendance_code))) %>% 
  mutate(tutoring_date = case_when(
    Day == "Monday" ~ as_date(week_of) + 0,
    Day == "Tuesday" ~ as_date(week_of) + 1,
    Day == "Wednesday" ~ as_date(week_of) + 2,
    Day == "Thursday" ~ as_date(week_of) + 3,
    Day == "Friday" ~ as_date(week_of) + 4), .before = "Day")%>% 
  relocate(week_of, .after = "Orientation")
```

{{< /panel >}}
{{< /panelset >}}

## Step 3: Visualize the data

Now that the data's in the proper format, it's time to visualize it. The Flexdashboard package has all sorts of neat functions like value boxes and gauges that are great for monitoring your metrics. If you want to visualize your data over time, the flexdashboard package integrates nicely with ggplot2 and plotly packages. The graph below shows attendance trends over time.

{{< panelset class="greetings" >}}
{{< panel name="Value Box :wave:" >}}

```{r eval=FALSE}

n_total_sessions <- sum(df_attendance_long$attendance_code == "1",  
                      df_attendance_long$attendance_code == "2",
                      df_attendance_long$attendance_code == "3", na.rm = TRUE)

valueBox(
  n_total_sessions,
  icon = "fa-pencil",
  href = gs_tutoring)
```

## Step 4: Throw in some datatables

Another fave feature in my dashboards are datatables. Datatables 

And VOILA..a Flexdashboard! 

I'm technically cheating by omitting a final and arguably the most important step which is publishing your dashboard. Following the steps above will get you a dashboard, but it will accessible only to you. When creating a dashboard, nine  times out of ten you'll want it to be publicly accessible to other people. 

Flexdashboard can be pushed to the web using Shiny, which I must admit is no simple feat and honestly deserves its own post which I'll save for another day. But if anyone is interested in trying out Flexdashboard, feel free to reach out to me. I'm happy to help out if I can.
